---
title: 'Coffee Future Price Pediction Project: ICFUT B3 Dataset'
subtitle: 'HarvardX - PH125.9x Data Science: Capstone Course'
author: "Mauricio Rabelo Soares"
date: "03 set 2022"
output:
  pdf_document:
      latex_engine: lualatex
  html_document:
    df_print: paged
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction

A Machine Learning for Factor Investing is a equity **investment strategies** that are built on **firm characteristics**, the goal of is this approach is to determine the model that maps the time-t characteristics of firms to their future performance.[^1] Factor investing is a subfield of a large discipline that encompasses asset allocation, quantitative trading and wealth management. Its premise is that differences in the returns of firms can be explained by the characteristics of these firms. In this project we use a similar approach but, instead a firm performance, our outcome is the 4/5 Arabica Coffee Futures contract performance traded at B3, which ticker is ICF.[^2] The goal of this project is to create a decision trade system, which result of the last 22 trade days is positive, or $r_m$ \> 0, using all the tools we have learn throughout the multi-part course in HarvardX's Data Science Professional Certificate series.

[^1]: <http://www.mlfactor.com/index.html>

[^2]: <https://www.b3.com.br/en_us/products-and-services/trading/commodities/product-sheet-8AE490C96D41D3A2016D46017EC97262.htm>

As presented in the Machine Learning course [\^2] [\^3], a recommendation systems for a movie use ***ratings*** **(*y~u,i~*)** that ***users*** **(*u*)** have given to some items, or ***movies*** **(*i*)**, to make specific recommendations to specific user. In this model, movies for which a high rating is predicted for a given user are then recommended to that user.

For this challenge we going to use a dataset provided by the Nelogica and Clear[^3]. The dataset begin in January 2003 and has **4468 decisions** applied to **1 asset** with **X features**.

[^3]: <https://corretora.clear.com.br/plataformas/profit-trader-clear/>

# Methods

The methods section explains the process and techniques used in this project

The **label** $\mathbf{y}=y_i$ of this project are approximated by functions of features $\mathbf{X}_i=(x_{_i,1},\dots,x_{i,K})$ . The dimension of the feature matrix $X$ is $I×K$: there are $I$ **instances**, **records**, or **observations** and each one of them has $K$ **features**, **inputs**, or **predictors** which will serve as **independent** and **explanatory** variables. We will write $x_i$ for one instance (one row) of $X$ or $x_k$ for one (feature) column vector of $X$.

The purpose of this project is to forecast **categories** (also called **classes**), "Buy," "Neutral" or "Sell." In order to be able to perform this type of classification analysis, we create additional labels that are categorical and continuous. The

Notation type pertains to finance We will work with daily returns $r_{t,n}=p_{t,n}/p_{t-1,n}-1$

computed from price data. Here t is the time index and n the asset index. Unless specified otherwise, the return is always computed over one period, though this period can sometimes be one month or one year. Whenever confusion might occur, we will specify other notations for returns.

**Evaluation metrics**

Avoiding False Positives: A Protocol

1. Research Motivation: ex ante/ex post economic foundation\
a) Does the model have a solid economic foundation?

Coffee It is one of the most important commodities traded globally and one of the most popular beverages in the world. World production is spread over 55 developing countries and is conducted by about 26 million producers, cultivated in tropical and subtropical areas.

The development of local organizations or instruments to help managing price risk, such as organized futures, in developing countries, like Brazil, is a important step to minimize the risk and challenges of small producers, since the larger and more liquids futures markets are located in the United States and Europe.

The Futures Price Volatility is

\
b) Did the economic foundation or hypothesis exist before the research was conducted?\

2\. Multiple Testing and Statistical Methods\
a) Did the researcher keep track of all models and variables that were tried (both successful and unsuccessful) and are the researchers aware of the multiple-testing issue?\
b) Is there a full accounting of all possible interaction variables if interaction variables are used?\
c) Did the researchers investigate all variables set out in the research agenda or did they cut the research as soon as they found a good model?\

3\. Data and Sample Choice\
a) Do the data chosen for examination make sense? And, if other data are available, does it make sense to exclude these data?\
b) Did the researchers take steps to ensure the integrity of the data?\
c) Do the data transformations, such as scaling, make sense? Were they selected in advance? And are the results robust to minor changes in these transformations?\
d) If outliers are excluded, are the exclusion rules reasonable?\
e) If the data are winsorized, was there a good reason to do it? Was the winsorization rule chosen before the research was started? Was only one winsorization rule tried (as opposed to many)?\

4\. Cross-Validation\
a) Are the researchers aware that true out-of-sample tests are only possible in live trading?\
b) Are steps in place to eliminate the risk of out-of-sample \"iterations\" (i.e., an in-sample model that is later modified to fit out-of-sample data)?\
c) Is the out-of-sample analysis representative of live trading? For example, are trading costs and data revisions taken into account?\

5\. Model Dynamics\
a) Is the model resilient to structural change and have the researchers taken steps to minimize the overfitting of the model dynamics?\
b) Does the analysis take into account the risk/likelihood of overcrowding in live trading?\
c) Do researchers take steps to minimize the tweaking of a live model?\

6\. Complexity\
a) Does the model avoid the curse of dimensionality?\
b) Have the researchers taken steps to produce the simplest practicable model specification?\
c) Is an attempt made to interpret the predictions of the machine learning model rather than using it as a black box?

\
7. Research Culture\
a) Does the research culture reward quality of the science rather than finding the winning strategy?\
b) Do the researchers and management understand that most tests will fail?\
c) Are expectations clear (that researchers should seek the truth not just something that works) when research is delegated?

Category #1: Research Motivation

Category #2: Multiple Testing and Statistical Methods
Keep track of: what is tried, combinations of variables,
Beware the parallel universe problem.

Category #3: Sample Choice and Data
Define the test sample ex ante.
Ensure data quality
Document choices in data transformations
Do not arbitrarily exclude outliers.
Select Winsorization level before constructing the model.

Category #4: Cross-Validation
Acknowledge out of sample is not really out of sample
Understand iterated out of sample is not out of sample.
Do not ignore trading costs and fees.

Category #5: Model Dynamics
Be aware of structural changes.
Acknowledge the Heisenberg Uncertainty Principle and overcrowding.
Refrain from tweaking the model.

Category #6: Model Complexity
Beware the curse of dimensionality.
Pursue simplicity and regularization.
Seek interpretable machine learning.

Category #7: Research Culture
Establish a research culture that rewards quality.
Be careful with delegated research

## The process and techniques used

Recommendation systems are particularly useful when an individual needs to choose an item from a potentially overwhelming number of items that a service may offer. The recommendations system build in this project are made

## Data cleaning

For this project we going to use a subset of dataset provided by

for the sake of **reproducibility**, we will illustrate the concepts we present with examples of implementation based on a single financial dataset available at <https://github.com/mrabelosoares/Coffee-Future-Price-Prediction>. This dataset comprises information on XX contracts listed in the B3. The time range starts in January 2003 and ends in August 2022. For each point in time, XX **characteristics** describe the decision in the sample. These attributes cover a wide range of topics:

**Price** (Open, High, Low, Close); **Volume**, **Time** (Date, Week of the year, Year), **Return**; **Status**; **Multiplier**; Adjust and Price.

```{r Load Library, message=FALSE, warning=FALSE, echo=TRUE}

# Load library


```

```{r Creat and clean the dataset, message=FALSE, warning=FALSE, echo=TRUE}
options(timeout=100)
#create tempfile and download


#read the file and give name to columns
ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))
movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

#create data frame movielens
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                           title = as.character(title),
                                           genres = as.character(genres))
movielens <- left_join(ratings, movies, by = "movieId")

#remove temporary files
rm(dl, ratings, movies)

```

## Data exploration and visualization

The exploration and visualization of the data provide insightful information about the users, the movies and ratings. The first 5 rows of the dataset that we use in this project is presented in table 1. The columns `movieId`, `userId` and `rating` are the variable of interest.

```{r The first 5 rows of the dataset, echo=TRUE, message=FALSE, warning=FALSE}
knitr::kable(head(movielens %>% as_tibble(),5),
             caption = "The first 5 rows of the dataset movielens",
             align = "cccccc",
             position = "h") %>%
  kable_styling(latex_options = "scale_down")
```

The variables, `movieId`, `userId` and `rating`, that is going to be used to build the model is presented as a matrix in the figure 1. The figure have the movies (`moveId`) in the x axis, the users in the y axis (`userId`), and the respective rating (`rating`). The matrix, extract from a sample of 50 users and 50 movies, provide some insights about the behavior of some users, the preference for some movies, and the sparsity of the matrix. The goal of this project is to fill the blank spaces with a rate.

```{r Matrix users x movies x rating, echo=TRUE, message=FALSE, warning=FALSE,out.width = "50%",fig.align='center'}
#matrix users x movies x rating
set.seed(1, sample.kind="Rounding")
users <- sample(unique(movielens$userId), 50)
movielens %>% filter(userId %in% users) %>% 
  select(userId, movieId, rating) %>%
  spread(movieId, rating) %>% select(sample(ncol(.), 50)) %>% 
  as.matrix() %>% t(.) %>%
  image.plot(1:50, 1:50,. , 
             nlevel=9,
             xlab="Movies", 
             ylab="Users", 
             main= "Matrix: users x movies x rating",
             out.width = "50%")
```

To see a potential flaw in the data we make a slice of the top 5 most rated movies and most active users. The unique users that provided ratings, the unique movies that were rated and the unique rating provided by a unique user to a unique movie, are presented to illustrate the possible rating matrix $users \times movies$ = $10677 \times 69878 = 746087406$ and the realized rating matrix $10000054$, or 1.34% of points of the matrix is filled. The extremes values confirm that some users are much more actives than others, the most active user rated more than 50% of the total unique movies, and some movies have been rated for more than 1/3 of the total unique users.

```{r Top 5 most rating movies and users, unique variables, echo=TRUE, message=FALSE, warning=FALSE, out.width = "60%",fig.align='center'}
# Unique users, movies, rating
p0 <- tableGrob(movielens %>% summarize(n_users = n_distinct(userId), 
                                  n_movies = n_distinct(movieId),
                                  n_rating = length(rating)))

# Top 5 movies
p1 <- movielens %>%
  group_by(title) %>%
  summarize(count = n()) %>%
  arrange(-count) %>%
  top_n(5, count) %>%
  ggplot(aes(count, reorder(title, count))) +
  geom_bar(color = "black", fill = "#999999", stat = "identity") +
  geom_text(aes(label=count), position=position_dodge(width=0.9), hjust=1.5) +
  xlab("Number of Ratings") +
  ylab("Movies") +
  theme_bw()

# Top 5 users
p2 <- movielens %>%
  group_by(userId) %>%
  summarize(count = n()) %>%
  arrange(-count) %>%
  top_n(5, count) %>%
  ggplot(aes(count, reorder(userId, count))) +
  geom_bar(color = "black", fill = "#999999", stat = "identity") +
  geom_text(aes(label=count), position=position_dodge(width=0.9), hjust=1.5) +
  xlab("Number of Ratings") +
  ylab("Users") +
  theme_bw()

#Top 5 most rating movies and users, unique variables
gridExtra::grid.arrange(p1,
               arrangeGrob (p0, p2, ncol = 2), 
               nrow = 2, 
               top = "Top 5 most rating movies and users, unique variables")

```

The dataset distribution presented trough histograms provide some insights about the general proprieties of the data. As showed in the slice before some movies get rated more than others, and some user are more active than others.

```{r distribution of movies and user, message=FALSE,echo=TRUE, warning=FALSE,out.width = "60%",fig.align='center'}

p3 <- movielens %>% 
  count(movieId) %>% 
  ggplot(aes(n)) + 
  geom_histogram(bins = 100, color = "black", fill="#999999") + 
  scale_x_log10() + 
  ggtitle("Movies")
p4 <- movielens %>% 
  count(userId) %>% 
  ggplot(aes(n)) + 
  geom_histogram(bins = 100, color = "black", fill="#999999") + 
  scale_x_log10() + 
  ggtitle("Users")
gridExtra::grid.arrange(p3, p4, nrow = 2)

```

## Modeling approach

The recommendation system is better as its error has decreased, for this project the error is the typical error we make when predicting a movie rating $\left( \hat{y}_{u,i} - y_{u,i} \right)$. The loss function used to evaluate the models is based on the residual mean squared error ($RMSE$) on a test set. The definition of $RMSE$ includes $N$ , the number of user/movie combination, and the sum occurring over all these combination. In this case the Loss function is: $$
\mbox{RMSE} = \sqrt{\frac{1}{N} \sum_{u,i}^{} \left( \hat{y}_{u,i} - y_{u,i} \right)^2 }
$$

```{r RMSE, message=FALSE, warning=FALSE, echo=TRUE}

RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings - predicted_ratings)^2))
}
```

The model which $RMSE = 0$ is a perfect model prediction, or without errors. For this project the reported $RMSE < 0.86490$ is the goal. The $RMSE > 1$ means our error is larger than one star, which means a bad model.

### Machine Learning

The machine learning decisions are based on algorithms build with data, so for this project the dataset XXXXX are going to be used to train and test the model. The train_set and test_set are build trough function `createDataPartition` as presented:

```{r Create data partition train_set and temp, message=FALSE, warning=FALSE, echo=TRUE,}



```

### Naive approach

The simplest model to recommend a movie to any user, is the model that predict the same rate $\mu$ for all movies regardless of user. In the naive approach the variation of the differences is random, and the independent error $\epsilon_{u,i}$ centered at 0. The model looks like:

$$
Y_{u,i} = \mu + \varepsilon_{u,i}
$$

The average of all ratings is the estimate that minimize the $RMSE$, and if we fill the blank cells in the matrix with the $\mu$ we obtain the $\varepsilon_{u,i}$.

```{r Model 1 - Naive_rmse, echo=TRUE, message=FALSE, warning=FALSE}
#Average of all ratings


# Model 1 - Naive_rmse

```

### Bias approach

```{r Lambda, echo=TRUE, message=FALSE, warning=FALSE, out.width = "70%",fig.align='center'}
# Model 4 Regularization and RMSE

```

The values of ${b}_i$ and ${b}_u$ that minimize the full model and the code to calculate the $RMSE$ is presented below.

```{r Model 4, echo=TRUE, message=FALSE, warning=FALSE, out.width = "70%",fig.align='center'}
#minimize the full model
lambdas <- seq(0, 10, 0.25)

```

Finally we achieve our goal, the $RMSE$ is lower than 0.86490. But can we do better?

### Matrix Factorization approach

A matrix Factorization is the last approach to solve recommendation system problem presented in this

```{r Model 5 Matrix Factorization using recosystem, echo=TRUE, message=FALSE, warning=FALSE, results = FALSE}
# Model 5 Matrix Factorization using recosystem

```

```{r Model 5 RMSE, echo=TRUE, message=FALSE, warning=FALSE}
# Model 5 RMSE

```

The model improved substantially and we achieve the best result so far, we have our choice to train and test the dataset.

# Results

This section presents the modeling results and discusses the model performance of the `movielens` data frame. The code below provided by HarvardX and create the `edx` and `validation` datasets that will be used to train and test our final algorithm.

```{r Final validation, echo=TRUE, message=FALSE, warning=FALSE, results = FALSE}

```

```{r Final prediction, echo=TRUE, message=FALSE, warning=FALSE}
# Final prediction

```

The final results of matrix factorization is a $RMSE$ of 0.78.

# Conclusion

The recommendation system, developed in this project, provide suggestions for items that are most pertinent to a particular user. The goal of this project was achieved and the $RMSE$ of matrix factorization of 0.78 is lower than our target 0.86490. A larger dataset could provide a even better model since we have more data to train, and is one of the limitations of this project. The addition of more variables could be a good outlook for future works since can capture more structures and latent factors in the data.

# References

Coqueret, Guillaume, and Tony Guida. 2020. *Machine learning for factor investing: R version*. Chapman; Hall/CRC.
